\textit{Supplementary materials for}: Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-11-15 paper/main.tex
complete draft incorporating edits to address all reviewer concerns (still needs final round of proofreading)

\newcommand{\dynamicsRandom}{S1}
\newcommand{\dynamicsAdaptive}{S2}
\newcommand{\accuracyByList}{S3}
\newcommand{\clusterCorrs}{S4}
\newcommand{\fingerprintsRandom}{S5}
\newcommand{\fingerprintsAdaptive}{S6}
\newcommand{\recallInit}{S7}
\newcommand{\fingerprintTrajectoryRandom}{S8}

@@

From a theoretical perspective, we are interested in several core questions
organized around the central theme of how structure in our experiences affect
how we remember \textit{those} experiences, and also how we remember
\textit{future} experiences (which may or may not exhibit similar structure).
For example, when we distill participants' experiences down to simple word
lists that vary (meaningfully) along just a few feature dimensions, are there
important differences in which dimenesions influence participants' memories? Or
are all features essentially ``equally'' influential? Further, are there
differences in how specific features influence participants' memories for
ongoing versus future experiences? Are there interaction effects between
different features, or do people appear to treat each feature independently?
And are there individual differences in how people organize their
memories, or in how people are influenced by our experimental manipulations? If
so, what are those differences and which aspects of memory do they affect?

@@

Each of six \textit{order manipulation} conditions used a different
feature-based sorting procedure to order words on early lists, where each
sorting procedure relied on one relevant feature dimension. All of the
irrelevant features varied freely across words on early lists, in that we did
not consider irrelevant features in ordering the early lists. However, we note
that some features were correlated---for example, some semantic categories of
words referred to objects that tended to be a particular size, which meant that
category and size were not fully independent. On late lists, the words were
always presented in a randomized order (chosen anew for each participant). In
all of the order manipulation conditions, we varied words' font colors and
onscreen locations, as in the feature rich condition.

@@

Following some of our prior work~\citep{HeusEtal18a, HeusEtal21, MannEtal22},
we use low-dimensional embeddings to help visualize how participants' memory
fingerprints change across lists
(Figs.~\ref{fig:fingerprint-trajectories}A,~\fingerprintTrajectoryRandom A). To
compute a shared embedding space across participants and experimental
conditions, we concatenated the full set of across-participant average
fingerprints (for all lists and experimental conditions) to create a large
matrix with number-of-lists (16) $\times$ number-of-conditions (10, encluding
the adaptive condition) rows and seven columns (one for each feature clustering
score, plus an additional temporal clustering score column). We used principal
components analysis to project the seven-dimensional observations into a
two-dimensional space (using the two principal components that explained the
most variance in the data). For two visualizations
(Figs.~\ref{fig:fingerprint-trajectories}B,~and~\fingerprintTrajectoryRandom B),
we computed an additional set of two-dimensional embeddings for the
\textit{average} fingerprints across lists within a given list grouping (i.e.,
early or late). For those visualizations, we averaged across the rows (for each
condition and group of lists) in the combined fingerprint matrix prior to
projecting it into the shared two-dimensional space. This yielded a single
two-dimensional coordinate for each \textit{list group} (in each condition),
rather than for each individual list. We used these embeddings solely for
visualization. All statistical tests were carried out in the original
(seven-dimensional) feature spaces.

@@

@@

\caption{\textbf{Recall dynamics in feature rich free recall (order
manipulation conditions).} \textbf{A.} Behavioral plots. \textbf{Left panels.}
The probabilities of initiating recall with each word are plotted as a function
of presentation position. \textbf{Middle panels.} The conditional probabilities
of recalling each word are plotted as a function of the relative position (Lag)
to the words recalled just-prior. \textbf{Right panels.} The overall
probabilities of recalling each word are plotted as a function of presentation
position. \textbf{All panels.} Error ribbons denote bootstrap-estimated 95\%
confidence intervals (calculated across participants). Top panels display the
recall dynamics for early (order manipulation) lists in each condition (color).
Bottom panels display the recall dynamics for late (randomly ordered) lists.
See Figures~\dynamicsRandom~and~\dynamicsAdaptive~for analogous plots for the
random and adaptive conditions. \textbf{B.} Proportion of event boundaries (see
\textit{Identifying event boundaries}) for each condition's feature of focus,
plotted as a function of presentation position.}

@@

We also looked for more subtle group-level patterns. For example, perhaps
sorting early lists by one feature dimension could affect how participants
cluster \textit{other} features (on early and/or late lists) as well. We
defined participants' \textit{memory fingerprints} as the set of their temporal
and feature clustering scores (see \textit{Memory fingerprints}). A
participant's memory fingerprint describes how they tend to retrieve memories
of the studied items, perhaps searching in parallel through several feature
spaces (or along several representational dimensions). To gain insights into
the dynamics of how participants' clustering scores tended to change over time,
we computed the average (across participants) fingerprint from each list, from
each order manipulation condition (Fig.~\ref{fig:fingerprint-trajectories}). We
projected these fingerprints into a two-dimensional space to help visualize the
dynamics (top panels; see \textit{Computing low-dimensional embeddings of
memory fingerprints}). We found that participants' average fingerprints tended
to remain relatively stable on early lists, and exhibited a ``jump'' to another
stable state on later lists. The sizes of these jumps varied somewhat across
conditions (the Euclidean distances between fingerprints in their original high
dimensional spaces are displayed in the bottom panels). We also averaged the
fingerprints across early and late lists, respectively, for each condition
(Fig.~\ref{fig:fingerprint-trajectories}B). We found that participants'
fingerprints on early lists seem to be influenced by the order manipulations
for those lists (see the locations of the circles in
Fig.~\ref{fig:fingerprint-trajectories}B). There also seemed to be some
consistency across different features within a broader type. For example, both
semantic feature conditions (category and size; purple markers) diverge in a
similar direction from the group; both lexicographic feature conditions (length
and first letter; yellow markers) diverge in a similar direction; and both
visual conditions (color and location; green) also diverge in a similar
direction. But on late lists, participants' fingerprints seem to return to a
common state that is roughly shared across conditions (i.e., the stars in that
panel are clumped together).

@@

@@

A large number of prior studies have compared participants' memories for
categorized word lists that are presented in blocked versus random orders. In
``blocked'' lists, all of the words from a given semantic category (e.g.,
animals) are presented consecutively, whereas in ``random'' lists, the words
from different categories are intermixed. Most of these studies report that
participants tend to better remember blocked (versus random)
lists~\citep{CofeEtal66, DAgo69, Dall64, Shap70, ShapPalm70, Kint70,
BoweEtal68, BoweEtal69, LuekEtal71, Puff74}. Other studies suggest that these order
effects may also be modulated by factors like list length and the numbers of
examplars in each cateogry~\citep[e.g.,][]{BorgMang72}.

>>>>

\newcommand{\dynamicsRandom}{S1}
\newcommand{\dynamicsAdaptive}{S2}
\newcommand{\accuracyByList}{S3}
\newcommand{\clusterCorrs}{S4}
\newcommand{\fingerprintsRandom}{S5}
\newcommand{\fingerprintsAdaptive}{S6}
\newcommand{\recallInit}{S7}
\newcommand{\fingerprintTrajectoryRandom}{S8}
\newcommand{\featureCorrs}{S9}

@@

From a theoretical perspective, we are interested in several core questions
organized around the central theme of how structure in our experiences affect
how we remember \textit{those} experiences, and also how we remember
\textit{future} experiences (which may or may not exhibit similar structure).
For example, when we distill participants' experiences down to simple word
lists that vary (meaningfully) along just a few feature dimensions, are there
important differences in which dimensions influence participants' memories? Or
are all features essentially ``equally'' influential? Further, are there
differences in how specific features influence participants' memories for
ongoing versus future experiences? Are there interaction effects between
different features, or do people appear to treat each feature independently?
And are there individual differences in how people organize their
memories, or in how people are influenced by our experimental manipulations? If
so, what are those differences and which aspects of memory do they affect?

@@

Each of six \textit{order manipulation} conditions used a different
feature-based sorting procedure to order words on early lists, where each
sorting procedure relied on one relevant feature dimension. All of the
irrelevant features varied freely across words on early lists, in that we did
not consider irrelevant features in ordering the early lists. However, we note
that some features were correlated---for example, some semantic categories of
words referred to objects that tended to be a particular size, which meant that
category and size were not fully independent (Fig.~\featureCorrs). On late
lists, the words were always presented in a randomized order (chosen anew for
each participant). In all of the order manipulation conditions, we varied
words' font colors and onscreen locations, as in the feature rich condition.

@@

Following some of our prior work~\citep{HeusEtal18a, HeusEtal21, MannEtal22},
we use low-dimensional embeddings to help visualize how participants' memory
fingerprints change across lists
(Figs.~\ref{fig:fingerprint-trajectories}A,~\fingerprintTrajectoryRandom A). To
compute a shared embedding space across participants and experimental
conditions, we concatenated the full set of across-participant average
fingerprints (for all lists and experimental conditions) to create a large
matrix with number-of-lists (16) $\times$ number-of-conditions (10, including
the adaptive condition) rows and seven columns (one for each feature clustering
score, plus an additional temporal clustering score column). We used principal
components analysis to project the seven-dimensional observations into a
two-dimensional space (using the two principal components that explained the
most variance in the data). For two visualizations
(Figs.~\ref{fig:fingerprint-trajectories}B,~and~\fingerprintTrajectoryRandom B),
we computed an additional set of two-dimensional embeddings for the
\textit{average} fingerprints across lists within a given list grouping (i.e.,
early or late). For those visualizations, we averaged across the rows (for each
condition and group of lists) in the combined fingerprint matrix prior to
projecting it into the shared two-dimensional space. This yielded a single
two-dimensional coordinate for each \textit{list group} (in each condition),
rather than for each individual list. We used these embeddings solely for
visualization. All statistical tests were carried out in the original
(seven-dimensional) feature spaces.

@@

A core assumption of our approach is that each participant organizes
their memories in a unique way. We defined each participant's \textit{memory
fingerprint} as the set of their permutation-corrected clustering scores across
all dimensions we tracked in our study, including their six feature-based
clustering scores (category, size, length, first letter, color, and location)
and their temporal clustering score. Conceptually, a participant's memory
fingerprint describes their tendency to order in their recall sequences (and,
presumably, organize in memory) the studied words along each dimension. If
these memory fingerprints are truly unique to each participant, then we would
expect that the estimated fingerprints computed for a given participant, on
different lists, should be more similar than the estimated fingerprints computed
for different participants. We reasoned that the feature rich condition would
provide the best opportunity to test this assumption, since the clustering
scores would not be potentially confounded by order manipulations. To test our
``unique memory fingerprint'' assumption, we compared the similarity
(correlation) between the fingerprint from a single list (from one participant)
and (a) the average fingerprint from all other lists from the same participant
versus (b) the average fingerprint from each other participant (across all of
their lists). We found that participants' fingerprints on a held-out list are
reliably more similar to the same participant's fingerprints on other lists
than to other participants' fingerprints ($t(70280) = 5.077, p < 0.001, d =
0.162, \mathrm{CI} = [3.086, 6.895]$). This suggests that participants'
fingerprints are stable across lists, and that each participant's fingerprint
is unique to them.

@@

\caption{\textbf{Recall dynamics in feature rich free recall (order
manipulation conditions).} \textbf{A.} Behavioral plots. \textbf{Left panels.}
The probabilities of initiating recall with each word are plotted as a function
of presentation position. \textbf{Middle panels.} The conditional probabilities
of recalling each word are plotted as a function of the relative position (Lag)
to the words recalled just-prior. \textbf{Right panels.} The overall
probabilities of recalling each word are plotted as a function of presentation
position. \textbf{All panels.} Error ribbons denote bootstrap-estimated 95\%
confidence intervals (calculated across participants). Top panels display the
recall dynamics for early (order manipulation) lists in each condition (color).
Bottom panels display the recall dynamics for late (randomly ordered) lists.
See Figures~\dynamicsRandom~and~\dynamicsAdaptive~for analogous plots for the
random and adaptive conditions. \textbf{B.} Feature distances ($z$-scored
within condition) between the features of successively presented words (see
\textit{Defining feature-based distances}), for each condition's feature of
focus, plotted as a function of presentation position.\textbf{C.} Proportion of
event boundaries (see \textit{Identifying event boundaries}) for each
condition's feature of focus, plotted as a function of presentation position.}

@@

We also looked for more subtle group-level patterns. For example, perhaps
sorting early lists by one feature dimension could affect how participants
cluster \textit{other} features (on early and/or late lists) as well. As
described above, a participant's memory fingerprint describes how they tend to
retrieve memories of the studied items, perhaps searching in parallel through
several feature spaces (or along several representational dimensions). To gain
insights into the dynamics of how participants' clustering scores tended to
change over time, we computed the average (across participants) fingerprint
from each list, from each order manipulation condition
(Fig.~\ref{fig:fingerprint-trajectories}). We projected these fingerprints into
a two-dimensional space to help visualize the dynamics (top panels; see
\textit{Computing low-dimensional embeddings of memory fingerprints}). We found
that participants' average fingerprints tended to remain relatively stable on
early lists, and exhibited a ``jump'' to another stable state on later lists.
The sizes of these jumps varied somewhat across conditions (the Euclidean
distances between fingerprints in their original high dimensional spaces are
displayed in the bottom panels). We also averaged the fingerprints across early
and late lists, respectively, for each condition
(Fig.~\ref{fig:fingerprint-trajectories}B). We found that participants'
fingerprints on early lists seem to be influenced by the order manipulations
for those lists (see the locations of the circles in
Fig.~\ref{fig:fingerprint-trajectories}B). There also seemed to be some
consistency across different features within a broader type. For example, both
semantic feature conditions (category and size; purple markers) diverge in a
similar direction from the group; both lexicographic feature conditions (length
and first letter; yellow markers) diverge in a similar direction; and both
visual conditions (color and location; green) also diverge in a similar
direction. But on late lists, participants' fingerprints seem to return to a
common state that is roughly shared across conditions (i.e., the stars in that
panel are clumped together).

@@

We note that the multivariate aspect of the adaptive condition (i.e., sorting
lists simultaneously along multiple feature dimensions) provides an important
contrast with the order order manipulation conditions, where we sort lists
along only a single feature dimension in each condition. We found that
participants ``naturally'' clustered their recalls along multiple feature
dimensions, even when the lists they studied were not sorted along those
dimensions (as in the feature rich condition). A caveat is that the
\textit{specific} feature dimensions participants tended to cluster along
varied across participants. One way to quantify the multidimensional nature of
participants' clustering tendencies is to sort each partipant's clustering
scores (for each of the six feature dimensions, along with a seventh dimension
to capture temporal clustering). We can then ask whether the distribution of
clustering scores at each ``rank'' within the sorted set of scores for each
participant has a mean that is reliably different from a chance value of 0.5.
We carried out these tests for each set of ranked scores, and found that
participants in the feature rich condition reliably cluster their recalls along
at least three dimensions, including temporal clustering (which was often
ranked highest); Rank 1: $t(66) = 12.751, p < 0.001, d = 0.162, \mathrm{CI} =
[8.702, 20.013]$; Rank 2: $t(66) = 8.196, p < 0.001, d = 0.162, \mathrm{CI} =
[4.794, 12.978]$; Rank 3: $t(66) = 3.243, p = 0.002, d = 0.162, \mathrm{CI} =
[1.028, 7.051]$; Rank 4: $t(66) = -3.112, p = 0.003, d = 0.162, \mathrm{CI} =
[-5.282, -1.920]$; Rank 5: $t(66) = -7.154, p < 0.001, d = 0.162, \mathrm{CI} =
[-12.649, -5.568]$; Rank 6: $t(66) = -12.608, p < 0.001, d = 0.162, \mathrm{CI}
= [-22.114, -9.347]$; Rank 7: $t(66) = -18.397, p < 0.001, d = 0.162,
\mathrm{CI} = [-27.238, -14.073]$.

@@

A large number of prior studies have compared participants' memories for
categorized word lists that are presented in blocked versus random orders. In
``blocked'' lists, all of the words from a given semantic category (e.g.,
animals) are presented consecutively, whereas in ``random'' lists, the words
from different categories are intermixed. Most of these studies report that
participants tend to better remember blocked (versus random)
lists~\citep{CofeEtal66, DAgo69, Dall64, Shap70, ShapPalm70, Kint70,
BoweEtal68, BoweEtal69, LuekEtal71, Puff74}. Other studies suggest that these order
effects may also be modulated by factors like list length and the numbers of
exemplars in each category~\citep[e.g.,][]{BorgMang72}.

### 2023-11-15 paper/supplement.tex
complete draft incorporating edits to address all reviewer concerns (still needs final round of proofreading)

>>>>

    \begin{figure}[tp] \centering
        \includegraphics[width=0.6\textwidth]{figures/feature_correlations}

        \caption{\textbf{Correlations between features.} Within each list, for each participant, we
        computed the set of distances between each pair of words, along each feature dimension.  We
        then combined these pairwise distances across all lists and participants, and computed the
        Spearman correlation between the distances for each pair of feature dimensions.  The correlation
        coefficients are displayed in the heatmap, annotated with each cell's value.} \label{fig:feature-correlations}
